\begin{thebibliography}{99}
\addcontentsline{toc}{chapter}{Bibliography}

\bibitem{AndersonAnderson}
Anderson, M., \& Anderson, S. L. (Eds.). (2011).
\textit{Machine Ethics}.
Cambridge University Press.

\bibitem{Anderson2018}
Anderson, J., Rainie, L., and Luchsinger, A. (2018).
\textit{Artificial Intelligence and the Future of Humans}.
Pew Research Center.

\bibitem{Allcott2020}
Allcott, Hunt, Braghieri, Luca, Eichmeyer, Sarah, and Gentzkow, Matthew (2020).
\textit{The welfare effects of social media}.
American Economic Review, 110(3), 629-76.

\bibitem{Auxier2021}
Auxier, Brooke, and Anderson, Monica (2021).
\textit{Social media use in 2021}.
Pew Research Center.

\bibitem{Allen2006}
Allen, Colin and Wallach, Wendell and Smit, Iva. (2006).
\textit{Why machine ethics?},
In: IEEE Intelligent Systems, 21(4), pp. 12--17.
IEEE.

\bibitem{Allen_Wallach2012}
Allen, C., \& Wallach, W. (2012).
\textit{Moral machines: contradiction in terms or abdication of human responsibility}.
In \textit{Robot ethics: The ethical and social implications of robotics} (pp. 55--68).
MIT Press Cambridge. Mass.

\bibitem{Aristotle}
Aristotle. (1984).
\textit{The Complete Works of Aristotle: The Revised Oxford Translation}.
Princeton University Press.

\bibitem{Bail2021}
Bail, Christopher A. (2021).
\textit{Breaking the social media prism: How to make our platforms less polarizing}.
Princeton University Press.

\bibitem{bandura1986}
Bandura, A. (1986).
\textit{Social foundations of thought and action}.
Englewood Cliffs, NJ, 1986.

\bibitem{Bryson}
Bryson, J. J. (2010).
\textit{Robots should be slaves}. 
In Close engagements with artificial companions: Key social, psychological, ethical and design issues (pp. 63-74). 
John Benjamins Publishing.

\bibitem{bird2000}
Bird, A. (2000).
\textit{Thomas Kuhn.}
Princeton University Press.

\bibitem{bricmont2016}
Bricmont, J. (2016).
\textit{Making Sense of Quantum Mechanics.}
Springer.

\bibitem{costa1992}
Costa, P. T., \& McCrae, R. R. (1992).
Four ways five factors are basic. 
\textit{Personality and individual differences}, 13(6), 653-665.

\bibitem{chalmers2013}
Chalmers, A. F. (2013).
\textit{What is this thing called science?}
Hackett Publishing.

\bibitem{laudan1987}
Laudan, L. (1987).
Progress or Rationality? The Prospects for Normative Naturalism.
\textit{American Philosophical Quarterly, 24(1), 19-31.}

\bibitem{woodward2007}
Woodward, J. (2007).
\textit{Making things happen: A theory of causal explanation.}
Oxford university press.


\bibitem{Dennett}
Dennett, D. C. (1971).
\textit{Intentional systems}.
The Journal of Philosophy, 68(4), 87-106.

\bibitem{Dwyer2020}
Dwyer, Ryan J., El-Bardicy, Mostafa, and Hakami, Tahani (2020).
\textit{Seeking and avoiding digital distractions in the workplace}.
Information Systems Journal, 30(5), 845-874.

\bibitem{Floridi}
Floridi, L. (2008).
\textit{Levels of Abstraction and the Foundation of Computational Ethics}.
APA Newsletter on Philosophy and Computers, 8(1), 3-5.

\bibitem{Goodfellow2016}
Goodfellow, I., Bengio, Y., and Courville, A. (2016).
\textit{Deep Learning}.
MIT Press.

\bibitem{greenwald2006}
Greenwald, A. G., \& Krieger, L. H. (2006).
Implicit bias: Scientific foundations.
\textit{California law review}, 94(4), 945-967.

\bibitem{Hampton}
Hampton, K. N., Sessions, L. F., Her, E. J., and Rainie, L. (2009). 
\textit{Social isolation and new technology}. 
Pew Internet and American Life Project.

\bibitem{IFR2019}
International Federation of Robotics (IFR). (2019).
\textit{World Robotics Report}.
IFR.

\bibitem{Mendelson}
Mendelson, E. (2009).
\textit{Introduction to mathematical logic}.
CRC Press.

\bibitem{Minsky1985} Minsky, M. (1985). \textit{The Society of Mind}. Simon and Schuster.

\bibitem{Moor}
Moor, J. H. (2006).
\textit{The nature, importance, and difficulty of machine ethics}.
IEEE intelligent systems, 21(4), 18-21.

\bibitem{Pantic2014a}
Pantic, I. (2014).
\textit{Online social networking and mental health},
Cyberpsychology, Behavior, and Social Networking,
volume 17, number 10,
Mary Ann Liebert Inc 140 Huguenot Street 3rd Floor New Rochelle NY 10801 USA.

\bibitem{Pantic2014b}
Pantic, Maja and Vinciarelli, Alessandro (2014),
\textit{Social signal processing},
The Oxford handbook of affective computing, page 84

\bibitem{Primack}
Primack, B. A., Shensa, A., Sidani, J. E., Whaite, E. O., Lin, L. Y., Rosen, D., Colditz, J. B., Radovic, A., and Miller, E. (2017). 
\textit{Social media use and perceived social isolation among young adults in the U.S.},
American Journal of Preventive Medicine, 53(1), 1-8. 
DOI: 10.1016/j.amepre.2017.01.010


\bibitem{Russell}
Russell, B. (1919).
\textit{Introduction to Mathematical Philosophy}.
London: George Allen \& Unwin.

\bibitem{RussellNorvig}
Russell, S., \& Norvig, P. (2016).
\textit{Artificial Intelligence: A Modern Approach}.
Malaysia; Pearson Education Limited.

\bibitem{Shoham2018}
Shoham, Y., Perrault, R., Brynjolfsson, E., Clark, J., Manyika, J., Niebles, J.C., Lyon, T., Etchemendy, J. (2018).
\textit{The AI Index 2018 Annual Report}.
AI Index Steering Committee, Human-Centered AI Initiative, Stanford University.

\bibitem{Silver2018}
Silver, D. et al. (2018).
\textit{A general reinforcement learning algorithm that masters chess, shogi, and Go through self-play}.
Science, 362(6419), 1140-1144.

\bibitem{Stone2016}
Stone, P. et al. (2016).
\textit{Artificial Intelligence and Life in 2030}.
One Hundred Year Study on Artificial Intelligence: Report of the 2015-2016 Study Panel, Stanford University.

\bibitem{Taylor}
Taylor, C. (1985).
\textit{Human Agency and Language: Philosophical Papers, Volume 1}.
Cambridge University Press.

\bibitem{Turkle}
Turkle, S. (2011).
\textit{Alone together: Why we expect more from technology and less from each other}.
Basic books.

\bibitem{Zermelo}
Zermelo, E. (1908).
\textit{Investigations in the foundations of set theory I}.
In From Kant to Hilbert: A Source Book in the Foundations of Mathematics, Ewald, W. (ed.), Oxford University Press.

\bibitem{kahneman2011}
Kahneman, D. (2011).
\textit{Thinking, fast and slow}.
Macmillan.

\bibitem{james1884}
James, W. (1884).
What is an Emotion?. 
\textit{Mind}, 9(34), 188-205.

\bibitem{Misra}
Misra, S., Cheng, L., Genevie, J., and Yuan, M. (2016).
\textit{The iPhone Effect: The Quality of In-Person Social Interactions in the Presence of Mobile Devices.} Environment and Behavior, 48(2), 275-298.

\bibitem{Turkle}
Turkle, S. (2011).
\textit{Alone Together: Why We Expect More from Technology and Less from Each Other}.
New York: Basic Books.

\bibitem{tversky1973}
Tversky, A., \& Kahneman, D. (1973).
Availability: A heuristic for judging frequency and probability.
\textit{Cognitive psychology}, 5(2), 207-232.

\bibitem{thaler2008}
Thaler, R. H., \& Sunstein, C. R. (2008).
\textit{Nudge: Improving decisions about health, wealth, and happiness}.
Yale University Press.

\bibitem{Vosoughi2018}
Vosoughi, Soroush, Roy, Deb, and Aral, Sinan (2018).
\textit{The spread of true and false news online}.
Science, 359(6380), 1146-1151.

\bibitem{Haidt2012}
Haidt, Jonathan (2012).
\textit{The Righteous Mind: Why Good People Are Divided by Politics and Religion}.
Pantheon

\bibitem{Xerxa2023}
Xerxa, Yllza and Rescorla, Leslie A and Shanahan, Lilly and Tiemeier, Henning and Copeland, William E., 	(2023)
\textit{Childhood loneliness as a specific risk factor for adult psychiatric disorders},
Psychological Medicine, Volume 53 number 1, pages 227--235,
Cambridge University Press.

\bibitem{Oda2015}
Oda, R., Kato, Y., \& Hiraishi, K. (2015).
\textit{The watching-eye effect on prosocial lying}.
Evolutionary Psychology, 13(3), 1474704915594959.
Los Angeles, CA: Sage Publications.

\bibitem{Atran2004}
Atran, S. \& Norenzayan, A. (2004).
\textit{Religion's Evolutionary Landscape: Counterintuition, Commitment, Compassion, Communion}.
Behavioral and Brain Sciences, 27(6), 713-770.

\bibitem{Bering2005}
Bering, J.M., McLeod, K., \& Shackelford, T.K. (2005).
\textit{Reasoning about dead agents reveals possible adaptive trends}.
Human Nature, 16(4), 360-381.

\bibitem{Shariff2007}
Shariff, A.F. \& Norenzayan, A. (2007).
\textit{God is watching you: Priming God concepts increases prosocial behavior in an anonymous economic game}.
Psychological Science, 18(9), 803-809.
Los Angeles, CA: SAGE Publications.

\bibitem{Sharkey2010}
Sharkey, A., \& Sharkey, N. (2010).
\textit{The crying shame of robot nannies: an ethical appraisal}.
Interaction Studies, 11(2), 161-190.

\bibitem{Vallor2016}
Vallor, S. (2016).
\textit{Technology and the virtues: A philosophical guide to a future worth wanting}.
Oxford: Oxford University Press.

\bibitem{Lin2012}
Lin, P., Abney, K., \& Bekey, G.A., eds. (2012).
\textit{Robot ethics: The ethical and social implications of robotics}.
Cambridge, MA: MIT Press.

\bibitem{Bryson2010}
Bryson, J.J. (2010).
\textit{Robots should be slaves}.
In Close engagements with artificial companions: Key social, psychological, ethical and design issues (pp. 63-74).
Amsterdam: John Benjamins Publishing Company.


\end{thebibliography}
